MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-04-30 MAESTROeX_I/algorithm.tex
paper updates
more updates

\section{Numerical Algorithm}

@@

In this step, using equation (\ref{eq:momentum}), 
we compute time-centered edge velocities, $\uadvonedag$, using
$\Ub^n$.  The $\dagger$ superscript refers to the
fact that the predicted velocity field does not satisfy the divergence
constraint.  We then construct $\uadvone$ from $\uadvonedag$
using a MAC projection.
We note that $\uadvone$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^n \uadvone\right) = \beta_0^n \left[S^{\nph,\star} - \frac{1}{\gammaonebar^np_0^n}\left(\frac{\partial p_0}{\partial t}\right)^{n-\myhalf} \right].\label{eq:div1}
\end{equation}
In practice, for all the projections in the algorithm, we solve a split dynamics
system for a base state velocity, and a perturbational velocity, so that we can properly
enforce boundary conditions at the edge of the star.
Refer to Appendix \ref{Sec:Projection} for details.

@@

The procedure to construct $\uadvtwodag$ is identical to the Godunov procedure
for computing $\uadvonedag$ in {\bf Step 3}, but uses
the updated value $S^{\nph}$ rather than $S^{\nph,\star}$.
We note that $\uadvtwo$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^{\nph} \uadvtwo\right) =
\beta_0^{\nph} \left[S^{\nph} - \frac{1}{\gammaonebar^{\nph}p_0^{\nph}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right],\label{eq:div2}
\end{equation}
where
\begin{equation}
\beta_0^{\nph} = \frac{ \beta_0^n +  \beta_0^{n+1,\star} }{2},
\qquad
\gammaonebar^{\nph} = \frac{ \gammaonebar^n +  \gammaonebar^{n+1,\star} }{2}.
\end{equation}

@@

Finally, we use an approximate nodal projection to define $\Ub^{n+1}$
from $\Ub^{n+1,\dagger},$  such that $\Ub^{n+1}$ approximately
satisfies the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
As part of the projection we also define the new-time perturbational pressure,
$\pi^\nph.$  This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.  Details of the approximate projection are given in Paper III.

>>>>

\section{Numerical Algorithm}\label{eq:algorithm}

@@

In this step, using equation (\ref{eq:momentum}), 
we compute time-centered edge velocities, $\uadvonedag$, using
$\Ub^n$.  The $\dagger$ superscript refers to the
fact that the predicted velocity field does not satisfy the divergence
constraint.  We then construct $\uadvone$ from $\uadvonedag$
using a MAC projection (see Appendix \ref{Sec:Projection} for details).
We note that $\uadvone$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^n \uadvone\right) = \beta_0^n \left[S^{\nph,\star} - \frac{1}{\gammaonebar^np_0^n}\left(\frac{\partial p_0}{\partial t}\right)^{n-\myhalf} \right].\label{eq:div1}
\end{equation}
In practice, for all the projections in the algorithm, we solve a split dynamics
system for a base state velocity, and a perturbational velocity, so that we can properly
enforce boundary conditions at the edge of the star.
Refer to Appendix \ref{Sec:Projection} for details.

@@

The procedure to construct $\uadvtwodag$ is identical to the Godunov procedure
for computing $\uadvonedag$ in {\bf Step 3}, but uses
the updated value $S^{\nph}$ rather than $S^{\nph,\star}$.
The $\dagger$ superscript refers to the
fact that the predicted velocity field does not satisfy the divergence
constraint.  We then construct $\uadvtwo$ from $\uadvtwodag$
using a MAC projection (see Appendix \ref{Sec:Projection} for details).
We note that $\uadvtwo$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^{\nph} \uadvtwo\right) =
\beta_0^{\nph} \left[S^{\nph} - \frac{1}{\gammaonebar^{\nph}p_0^{\nph}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right],\label{eq:div2}
\end{equation}
where
\begin{equation}
\beta_0^{\nph} = \frac{ \beta_0^n +  \beta_0^{n+1,\star} }{2},
\qquad
\gammaonebar^{\nph} = \frac{ \gammaonebar^n +  \gammaonebar^{n+1,\star} }{2}.
\end{equation}

@@

Finally, we use an approximate projection to define $\Ub^{n+1}$
from $\Ub^{n+1,\dagger},$  such that $\Ub^{n+1}$ approximately
satisfies the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
As part of the projection we also define the new-time perturbational pressure, $\pi^\nph$.
This pressure $\pi$ is a nodal quantity, and thus this projection step is often referred to as a nodal projection.
This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.

Refer to Appendix \ref{Sec:Projection} for more details about the projection steps.

### 2019-04-30 MAESTROeX_I/appendix.tex
more ferences
more updates

\appendix
\section{Projection Details}\label{Sec:Projection}
To enforce the divergence constraint in {\bf Step 3, 7}, and {\bf 11}, we use a projection method analgous to the methods originally developed for incompressible flow \citep{almgren1998conservative,bell1989second}.

Based on our past experience in the MAESTRO project, we have found it useful to split the velocity dynamics into a perturbational and base state velocity,
\begin{equation}
\Ub = \Ubt(\xb,t) + w_0(r,t)\eb_r,
\end{equation}
solve for each term separately, and immediately combine them to find a full velocity that satisfies the constraint.  We take that approach here, primarily because it allows us to enforce a boundary condition on $w_0$ at the edge of the star (i.e., the cutoff density location where we hold density constant).  Namely, to enforce that $r^2 w_0$ remain constant is difficult to do when solving for the full velocity.  So in practice, we solve the constraint over the lateral average,
\begin{equation}
\nabla\cdot\left(\beta_0\Ubt\right) = \beta_0(S-\overline{S})
\end{equation}
and sepratately solve for $w_0$ using 
\begin{equation}
\nabla\cdot(\beta_0w_0\eb_r) = \beta_0\left(\overline{S} - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).
\end{equation}
To find $w_0$ we integrate in 1D using the procedure in Appendix B of Paper V.

Note that in this approach, we estimated the time-derivative of the pressure in part by examining how laterally averaged $\rho'=\rho-\rho_0$ changed over time (quantified by $\eta_\rho = \overline{\rho'\Ub\cdot\eb_r}$.  After evolving the species, we compute this term in the same way as reported in Paper V for use in computing $w_0$.

\MarginPar{mention alt energy fix for projection?  also mention changes in $w_0$ computation for irregular base state}

>>>>

\appendix
\section{Projection Details}\label{Sec:Projection}
To enforce the divergence constraint in {\bf Step 3, 7}, and {\bf 11}, we use a projection method analgous to the methods originally developed for incompressible flow \citep{almgren1998conservative,bell1989second}.
The basic idea is to decompose the velocity field into a divergence-free component and a curl-free (gradient of a scalar field) component by solving a variable-coefficient Poisson equation for the scalar field.
The details for the MAC projection in {\bf Step 3} and {\bf Step 7} are given in Appendix B of Paper III.  The details of the nodal projection in {\bf Step 11} are given in Section 3.2 of Paper III.
We note that in the nodal projection, the gradient of the scalar field is used to update the perturbational pressure, $\pi$.

Based on our past experience in the MAESTRO project, we have found it useful to split the velocity dynamics into a perturbational and base state velocity,
\begin{equation}
\Ub = \Ubt(\xb,t) + w_0(r,t)\eb_r,
\end{equation}
solve for each term separately, and immediately combine them to find a full velocity that satisfies the constraint.  We take that approach here, primarily because it allows us to enforce a boundary condition on $w_0$ at the edge of the star (i.e., the cutoff density location where we hold density constant).  Namely, to enforce that $r^2 w_0$ remain constant is difficult to do when solving for the full velocity.  So in practice, we solve the constraint over the lateral average,
\begin{equation}
\nabla\cdot\left(\beta_0\Ubt\right) = \beta_0(S-\overline{S})
\end{equation}
and sepratately solve for $w_0$ using 
\begin{equation}
\nabla\cdot(\beta_0w_0\eb_r) = \beta_0\left(\overline{S} - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).
\end{equation}
To solve for $\Ubt$ we use a projection method, which involves the solution of a variable-coefficient Poisson solver to extract the curl-free component of the unprojected velocity.
Note that MAESTRO contains alternate low Mach number formulations that conserve total energy in stratified systems, with minimal changes to the code (see Appendix A of \cite{subChandra_II} for details).
To find $w_0$ we integrate in 1D using the procedure in Appendix B of Paper V.
Note that in this approach, we estimated the time-derivative of the pressure in part by examining how laterally averaged $\rho'=\rho-\rho_0$ changed over time (quantified by $\eta_\rho = \overline{\rho'\Ub\cdot\eb_r}$).  After evolving the species, we compute this term in the same way as reported in Paper V for use in computing $w_0$.
\MarginPar{mention changes in $w_0$ computation for irregular base state. Add graphic of temperature falling due to incorrect splitting?}
